### 2016-05-18 thesis_paper/source.tex
Thesis paper changes

\documentclass[10pt,journal,final,letterpaper,twocolumn]{IEEEtran}
%\documentclass{ssdbm}
\usepackage{wrapfig,subfigure,setspace,verbatim,cite,url,graphicx,float}
\usepackage{algorithmic,amsfonts,amssymb,amsmath,calc,calrsfs,cleveref}
\usepackage[justification=centering]{caption}
%\usepackage{fixltx2e}

\restylefloat{table}

% This was from the NEW paper
%----------------------------------------------------------
\RequirePackage{fix-cm}

\DeclareMathOperator{\var}{var}
\newcounter{qcounter}

%\smartqed  % flush right qed marks, e.g. at end of proof
%----------------------------------------------------------

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\newtheorem{remark}{Remark}

\newcommand{\SEC}[1] {Section \ref{#1}}
\newcommand{\Eq}[1] {Equation \ref{#1}}
\newcommand{\eq}[1] {Eq. (\ref{#1})}

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

@@

% paper title
\title{Parallelization of Push-based System for Molecular Simulation Data Analysis with GPU}

% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs

\author{Iliiazbek~Akhmedov, Yi-Cheng~Tu~$\dagger$, Vladimir Grupcev, Joseph Fogarty, Sagar Pandit % <-this % stops a space

%\thanks{* These authors contributed equally to this work.}
\thanks{$\dagger$ Author to whom all correspondence should be sent.}
\thanks{Iliiazbek Akhmedov, Vladimir Grupcev, and Yi-Cheng Tu are with the Department of Computer Science and Engineering, University of South Florida, 4202 E. Fowler Ave., ENB 118, Tampa, FL 33620, U.S.A. Emails:
akhmedovi@mail.usf.edu, ytu@cse.usf.edu}
\thanks{Joseph Fogarty and Sagar Pandit are with the Department of Physics,
University of South Florida, 4202 E. Fowler Ave., ISA 2019, Tampa,
FL 33620, U.S.A. Emails: jcfogart@mail.usf.edu, pandit@usf.edu} }

%\pubid{0000--0000/00\$00.00~\copyright~2002 IEEE}

% use only for invited papers
%\specialpapernotice{(Invited Paper)}

% make the title area
\maketitle

\begin{abstract}

Modern simulation systems generate big amount of data, which consequently has to be analyzed in a timely fashion. Traditional database management systems follow principle of pulling the needed data, processing it, and then returning the results. This approach is then optimized by means of caching, storing in different structures, or doing some sacrifices on precision of the results to make it faster. When it comes to the point of doing various queries that require analysis  of the whole data,  this design has the following disadvantages: considerable overhead on random I/O while reading from the simulation output files and low throughput of the data that consequently results in long latency, and, if there was any indexing to optimize selections, overhead of storing those becomes too big, too.

There is a new approach to this problem presented in the previous paper -- Push-based System for Molecular Simulation Data Analysis for processing network of queries proposed in the previous paper and its primary steps are: i) it uses traditional scan-based I/O framework to load the data from files to the main memory and then ii) the data is pushed through a network of queries which consequently filter the data and collect all the needed information which increases efficiency and data throughput. It has a considerable advantage in analysis of molecular simulation data, because it normally involves all the data sets to be processed by the queries.

In this paper, we propose improved version of Push-based System for Molecular Simulation Data Analysis. Its major difference with the previous design is usage of GPU for the actual processing part of the data flow. Using the same scan-based I/O framework the data is pushed through the network of queries which are processed by GPU, and due to the nature of science simulation data, this gives a big advantage for processing it faster and easier. In the old approach there were some custom data structures such as quad-tree for calculation of histograms to make the processing faster and those involved loss of data and some expectations from the data nature, too. In the new approach due to high performance of GPU processing and its nature, custom data structures were not even needed much though it didn't bear any loss in precision and performance.

\end{abstract}

\begin{IEEEkeywords}
Push-based system, molecular simulation, scientific databases,
spatial distance histogram, GPU, parallel processing, CUDA.
\end{IEEEkeywords}
%\vspace{-5mm}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}\label{sc:intro}

In various sciences simulation systems take big place and often times they may be the clue for results. One of such sciences, which is primarily related to this paper, is physics. In this case, from computer engineering point of view, simulations have the following flow:

@@

{\small
\begin{enumerate}
	\item initial physical properties are given to simulation software as arguments which are interpreted and provided in a language defined by the simulation software
	\item then it runs for given amount of time that can generally last from milliseconds to months or even more
	\item finally, the file generated during the simulation is analyzed to extract useful information
\end{enumerate}
}
\vspace{3mm}
The purpose of this paper is primarily focused on the third step, which basically involves the entire simulation data to be processed by the analysis software.

Big data processing is becoming one of the key issues with the amount of data being generated by modern systems. Eventually, this data is required to be processed on the fly, thus, analysis software should be able to handle massive data in a very short period of time. When working with huge volume of data, there are non-trivial issues arise. For example, it can take days and weeks to analyze big enough data sets, because the data cannot be simply loaded into memory, since it can get up to terabytes, thus, it has extra overhead because of widely used random access disk I/O framework to read from disk chunks by chunks. Besides it, analysis of the data can get even more complicated, since going through certain parts of the data once would not be enough, which leads to low throughput and efficiency of loaded data. One such example, the data analysis approach has polynomial complexity just for reading the data in order to come up with result, and since the data can't simply be saved in memory, it raises the overhead of disk I/O, too. Pull-based architectures in data processing engines are inefficient, since having a set of specific queries, in order to compute them all, it is needed to fetch data, filter it, and apply needed formulas. It is inefficient, since for every query the same chunk of data needs to be pulled into the memory at least the number of queries times or more in case of more complex queries.

As it has already been mentioned in the previous paper \cite{mainPaper}, one of the modern issues of analyzing massive data on the fly is social networks. .
"In order for a system to be able to perform analytical examination of the data produced in such streaming media, the system should have the capability of fast data access. The reason, the millions of data records(tweets) produced
every second. Moreover, these tweets may have
different geographical origin, introducing different languages and
forms and often times containing unsolicited messages, errors,
malicious content, etc. Therefore, some low level data uniformity
and cleaning on top of the data access and management issues should
be considered and possibly incorporated in the process of analytical
investigation in order to achieve relevant result. "\cite{mainPaper, nature_bigdata08,nature_bigdata12,science_social10,jcs_twitter11}

The primary focus and problem in this paper is scientific data analysis. Particles simulation is one of the most popular methods of analyzing certain chemical reactions, physical processes, or other behavior of different materials.
Molecular simulations (Molecular Dynamics) are applied in different fields and represent a method of analyzing physical movements of particles, atoms, and molecules in a fixed space with a given period of time, apparently with a possibility of giving initial state for each item that is involved in the process and can affect the system. This system is an N-body simulation. The number of atoms in simulations vary in hundred of thousands, particularly, we may observe two simulation systems of a collagen fiber structure and dipalmitoylphosphatidylcholine (DPPC) bi-layer lipid system consisting of 890,000 and 402,400 atoms respectively on Figure \ref{fg:collagen_dipal}. Simulation data represents number of records of physical properties such as mass, charge, velocity, coordinates, and forces for each item aggregated as frames, where each frame represents a snapshot of time, placed with a fixed time interval which may also vary depending on the simulation itself and simulation precision requirement. "Quantities measured during the simulations are analyzed to test the theoretical
model~\cite{Frenkel:api01,Landau:cup05}. In short, the MS is proven
and powerful tool for understanding the inner-workings of a
biological system, by supplying a model description of the
biophysical and biochemical processes that are being unfold at a
nanoscopic scale." \cite{mainPaper}

@@

In this paper, we are incorporating parallelization into the processing part of the proposed design by means of CUDA programming language for GPU. Since storage of the simulation data is very expensive, it might come to the point of analyzing the data on the fly (meaning running the simulation and analyzing it at the same time in a streaming manner), which leads to a problem of optimizing the processing part of the design proposed in the previous paper, because time spent on generating data should be tried to conceal the time spent on the processing part by means of overlapping or simply running it in a quick manner.

\begin{figure}
 \centerline{ \includegraphics[width=1\columnwidth]{images/sample_snapshot_of_simulation.eps} }
 \caption{ Snapshots of two MS systems: a collagen fiber structure with 890,000 atoms (top) and a dipalmitoylphosphatidylcholine (DPPC) bi-layer lipid system with 402,400 atoms (bottom) \cite{mainArticle}}
 \label{fg:collagen_dipal}
\end{figure}

\subsection{Problem Statement}
Simulation software systems, in general, follow the same methodology of running and storing simulation data. The simulation software system examples are: Gromacs \cite{GROMACS4}, VMD\cite{VMD}, MDAnalysis~\cite{MDAnalysis}, Wordom~\cite{wordom}, MD-TRACKS~\cite{MDtracks}, SimulaidOne~\cite{Simulaid}, Charmm~\cite{CHARMM}. In the type of simulations brought up as examples above, the flow of the data is the following. Once the simulation is run, the output files are contained as trajectory files with descriptors (they contain information about space dimensions, number of atoms and frames, etc.) that can be easily transposed into simple flat files containing the physical properties atom by atom, frame by frame, which are consequently read and processed by the proposed push-based system. Since we have certain amount of queries needed to be run on given simulation data, generating high I/O traffic followed by design of pull-based system is not considerable. The approach proposed in the previous paper is very good in terms of performance in comparison with the original pull-based system\cite{mainPaper}. The problem is still that some of the queries processed by those means are still improvable, especially taking into consideration the fact that in the used previous works for calculation of 2-body functions, which take the biggest time for processing~\cite{ytu:icde09, EDBT12}, we might have some error bounds, which might be unacceptable for certain simulation analysis where sacrifice on loss of data is unbearable. Besides it, although we might have huge performance boost on specific data structure as density map, we still have certain expectation from data nature (such as its uniformity), thus, it makes sense to have desire to simplify the processing and still having pure computation based on any values, considering that the queries should be available to be pre-programmed by user in a separate module of the code.

\begin{figure}
 \centerline{ \includegraphics[width=1\columnwidth]{images/speedup838K-pomalo-za-4.eps} }
 \caption{ Speed up over different levels of atom selection for 838,400 atoms \cite{mainPaper}}
 \label{fg:sample_estimation_old_paper}
\end{figure}

\subsection{Our approach with improvement}

Reading the data from files is a huge overhead, thus, speedup of push-based system in general is significant over pull-based system in our given case. 
It's been demonstrated in the previous paper with different amount of atoms, frames, and workloads. Since we have to load the data every time there is a different query versus push-based system loads a chunk of data once, it is quite noticeable how the framework contributes to efficiency of analysis of simulation data. You can observe some estimation samples in Figure \ref{fg:sample_estimation_old_paper}. Even though the memory loading and pushing steps are the key features of the proposed design, nevertheless, having general knowledge about the network of queries, in this paper, we will try to focus on optimizing the actual execution of them.

Some of the queries may be very slow due to their nature on a sequential type of computation,  especially, 2-body functions. Since the nature of the data that comes with simulation is basically physical properties of atoms, there is a lot of computation that involves independent primitive mathematical operations, which makes it a perfect problem for parallelism.

Although the original idea of push-based system for molecular simulation data analysis was focused on optimizing throughput and usage of loaded data, there were some extra approaches specifically for 2-body functions. For example, Density Map for Spacial Distance Histogram was used in order to avoid additional memory allocation and latency reduction with proven error bounds. SDH is a quite intensive and computational problem, especially with increasing amount of atoms.

@@

\subsection{Contribution and roadmap of the paper} 
The proposed improved version of push-based system for molecular simulation data analysis, we believe, gives an opportunity for scientists to run their analysis even faster now. Taking an advantage of GPU devices nature and incorporating parallelism and streaming for different types of queries, we have come up with a good speed up over sequential processing, which already had a good speed up in terms of performance in comparison with original pull-based approach. Since in this paper we don't expect certain form of data, we have come up with a tool that generates mock data for any number of atoms and frames, which can be used for performance tests that were done consequently, too. This mock data has exactly the same information that would come with Gromacs simulation files, and it has exactly the same format as in the previous paper \cite{mainPaper} right before loading it to memory. This improved version has been developed on Amazon Elastic GPU \cite{amazonGPU} nodes that consequently can be replicated and scaled up becoming more of a streaming distributed processing engine (which can be very effective on costs, since it is always easy to scale them up, shut them down and spin them back up), as well as on a simple desktop computer with a GPU device.
In general, major contribution of this paper is the framework of push-based system with an incorporated parallelism based on CUDA programming language. Besides it, there was developed a tool which generates mockup data of the same format as of the real MS simulations and represent a python script with arguments given as number of frames and atoms.

The structure of the framework which will be described later is developed in such a way that one can easily add new queries based on their complexity or modify existing ones adding appropriate selections and filters.

@@

\vspace{3mm}
\section{Related Work}\label{sc:relatedwork}
\vspace{3mm}

@@

\vspace{3mm}
\section{The system design}\label{sc:system}
\vspace{3mm}

@@

\subsection{Data flow architecture}

>>>>

\documentclass[11pt,a4paper]{report}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{microtype}
\usepackage[dvips]{graphicx}
\usepackage{xcolor}
\usepackage{times}
\usepackage{ragged2e}
\usepackage{titlesec}
\usepackage{wrapfig,subfigure,setspace,verbatim,cite,url,graphicx,float}
\usepackage{algorithmic,amsfonts,amssymb,amsmath,calc,calrsfs,cleveref}
\usepackage[justification=centering]{caption}
\usepackage{url}

\usepackage[
breaklinks=true,colorlinks=true,
%linkcolor=blue,urlcolor=blue,citecolor=blue,% PDF VIEW
linkcolor=black,urlcolor=black,citecolor=black,% PRINT
bookmarks=true,bookmarksopenlevel=2]{hyperref}

\usepackage{geometry}
% PDF VIEW
\geometry{total={210mm,297mm},
left=25mm,right=25mm,%
bindingoffset=0mm, top=25mm,bottom=25mm}
%PRINT
\geometry{total={210mm,297mm},
left=20mm,right=20mm,
bindingoffset=0mm, top=15mm,bottom=35mm}

\linespread{1.5}

\titleformat{\chapter}
{\normalfont\LARGE\bfseries}{\thechapter.}{1em}{}

%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%

@@

%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%
%   TITLEPAGE
%
%   due to variety of titlepage schemes it is probably better to make titlepage manually
%
%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%
\thispagestyle{empty}

{%%%
\sffamily
\centering

{\Large 
Parallelization of Push-based System for Molecular Simulation Data Analysis with GPU
}

\vspace{1.5cm}

{
by
}
\vspace{.5cm}

{ 
Iliiazbek Akhmedov
}

\vspace{3.5cm}

A thesis submitted in partial fulfillment

of the requirements for the degree of

Master of Science in Computer Science

Department of Computer Science and Engineering

College of Engineering

University of South Florida

\vspace{3.5cm}
Major Professor:  Yicheng Tu, Ph.D.
\vspace{\fill}

\vspace{.5cm}
Date of Approval:

N/A
\vspace{\fill}

Copyright \textcopyright\space 2016, Iliiazbek Akhmedov

%%%
}%%%

\newgeometry{left=20mm,right=20mm,top=35mm,bottom=35mm}

\pagestyle{plain}
\centering

\textbf{DEDICATION}
\vspace{0.5cm}

\begin{flushleft}
\hspace{3em} I dedicate my thesis work to my family who have always been supporting me throughout my life. I also dedicate it to my closest friends who have shared my passion and encouraged me to explore and try my ideas in life.
\end{flushleft}

\pagestyle{plain}
\centering
\textbf{ACKNOWLEDGMENTS}
\vspace{0.5cm}

\begin{flushleft}
\hspace{3em} I would like to thank my major professor throughout my master thesis Dr. Yicheng Tu for his guidance and knowledge that I received through all the time I worked with him. I really appreciate his contribution and very happy that I got a chance to work with him. I as well want to thank the committee members for their support and encouragement.
\end{flushleft}

\clearpage
\restoregeometry

\tableofcontents

\begin{abstract}
\justify
\hspace{3em} Modern simulation systems generate big amount of data, which consequently has to be analyzed in a timely fashion. Traditional database management systems follow principle of pulling the needed data, processing it, and then returning the results. This approach is then optimized by means of caching, storing in different structures, or doing some sacrifices on precision of the results to make it faster. When it comes to the point of doing various queries that require analysis  of the whole data,  this design has the following disadvantages: considerable overhead on traditional disk random I/O framework while reading from the simulation output files and low throughput of the data that consequently results in long latency, and, if there was any indexing to optimize selections, overhead of storing those becomes too big, too. Besides it, indexing will also cause delay during write operations and since most of the queries work with the entire data sets, indexing loses its point.

There is a new approach to this problem -- Push-based System for Molecular Simulation Data Analysis for processing network of queries proposed in the previous paper and its primary steps are: i) it uses traditional scan-based I/O framework to load the data from files to the main memory and then ii) the data is pushed through a network of queries which consequently filters the data and collect all the needed information which increases efficiency and data throughput. It has a considerable advantage in analysis of molecular simulation data, because it normally involves all the data sets to be processed by the queries.

In this paper, we propose improved version of Push-based System for Molecular Simulation Data Analysis. Its major difference with the previous design is usage of GPU for the actual processing part of the data flow. Using the same scan-based I/O framework the data is pushed through the network of queries which are processed by GPU, and due to the nature of science simulation data, this gives a big advantage for processing it faster and easier (it will be explained more in later sections). In the old approach there were some custom data structures such as quad-tree for calculation of histograms to make the processing faster and those involved loss of data and some expectations from the data nature, too. In the new approach due to high performance of GPU processing and its nature, custom data structures were not even needed much, though it didn't bear any loss in precision and performance.
\end{abstract}

%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%
%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%

\chapter{Introduction}

\justify
In various sciences simulation systems take big place and often times they may be the clue for results. One of such sciences, which is primarily related to this paper, is physics. In this case, from computer engineering point of view, simulations have the following flow:

@@

\begin{enumerate}
	\item initial physical properties are given to simulation software as arguments which are interpreted and provided in a language defined by the simulation software
	\item then it runs for given amount of time that can generally last from milliseconds to months or even more
	\item finally, the file generated during the simulation is analyzed to extract useful information
\end{enumerate}

\vspace{3mm}
The purpose of this paper is primarily focused on the third step, which basically involves the entire simulation data to be processed by the analysis software. It is important to underline scale of data being discussed. Molecular simulations represent systems starting from hundreds thousands of atoms and more which are tracked by frames as snapshot of specific time and can last for months. Thus, the size of data starts from couple of gigabytes up to terabytes and petabytes.

Big data processing is becoming one of the key issues with the amount of data being generated by modern systems. Eventually, this data is required to be processed on the fly, thus, analysis software should be able to handle massive data in a very short period of time. When working with huge volume of data, there are non-trivial issues arise. For example, it can take days and weeks to analyze big enough data sets, because the data cannot be simply loaded into memory, since it can get up to terabytes, thus, it has extra overhead because of traditional random access disk I/O framework to read from disk chunks by chunks. Besides it, analysis of the data can get even more complicated, since going through certain parts of the data once would not be enough, which leads to low throughput and efficiency of loaded data. One such example, the data analysis approach has polynomial complexity just for reading the data in order to come up with result, and since the data can't simply be saved in memory, it raises the overhead of disk I/O, too. Pull-based architectures in data processing engines are inefficient, since having a set of specific queries, in order to compute them all, it is needed to fetch data, filter it, and apply needed formulas. It is inefficient, since for every query the same chunk of data needs to be pulled into the memory at least the number of queries times or more in case of more complex queries.

As it has already been mentioned in the previous paper \cite{mainPaper}, one of the modern issues of analyzing massive data on the fly is social networks. "In order for a system to be able to perform analytical examination of the data produced in such streaming media, the system should have the capability of fast data access. The reason is millions of data records (tweets) produced every second. Moreover, these tweets may have different geographical origin, introducing different languages and forms and often times containing unsolicited messages, errors,
malicious content, etc. Therefore, some low level data uniformity
and cleaning on top of the data access and management issues should
be considered and possibly incorporated in the process of analytic
investigation in order to achieve relevant result".\cite{mainPaper, nature_bigdata08,nature_bigdata12,science_social10,jcs_twitter11} Another good example would be data scale of Facebook Inc. Being the largest social network, it has huge data coming in daily \cite{fbStats}. The data warehouse of Facebook is measured to be 300PB in 2014, and the daily growth is 600TB, most of the raw data is logs, which, most of the times, are used for analytics. Analytical queries involve entire data sets that will be handy for any future strategical planning in big companies, being able to observe data behavior and much more.

The primary focus and problem in this paper is scientific data analysis. Examples brought up above are real life problems that are very similar to the scientific problem that we are focusing on in terms of data sizes. Particles simulation is one of the most popular methods of analyzing certain chemical reactions, physical processes, or other behavior of different materials.
Molecular simulations (Molecular Dynamics) are applied in different fields and represent a method of analyzing physical movements of particles, atoms, and molecules in a fixed space with a given period of time, apparently with a possibility of giving initial state for each item that is involved in the process and can affect the system. These systems are most of the time N-body simulations. The number of atoms in simulations vary in hundred of thousands, particularly, we may observe two simulation systems of a collagen fiber structure and dipalmitoylphosphatidylcholine (DPPC) bi-layer lipid system consisting of 890,000 and 402,400 atoms respectively on Figure \ref{fg:collagen_dipal}. Simulation data represents number of records of physical properties such as mass, charge, velocity, coordinates, and forces for each item aggregated as frames, where each frame represents a snapshot of time, placed with a fixed time interval which may also vary depending on the simulation itself and simulation precision requirement. "Quantities measured during the simulations are analyzed to test the theoretical
model~\cite{Frenkel:api01,Landau:cup05}. In short, the MS is proven
and powerful tool for understanding the inner-workings of a
biological system, by supplying a model description of the
biophysical and biochemical processes that are being unfold at a
nanoscopic scale." \cite{mainPaper}

@@

In this paper, we are incorporating parallelization into the processing part of the proposed design by means of CUDA programming language for GPU. Since storage of the simulation data is very expensive, it might come to the point of analyzing the data on the fly (meaning running the simulation and analyzing it at the same time in a streaming manner), which leads to a problem of optimizing the processing part of the design proposed in the previous paper, because time spent on generating data should be tried to conceal the time spent on the processing part by means of overlapping or simply running it in a quick manner. We will talk about this in details as well in further sections.

\begin{figure}
 \centerline{ \includegraphics[width=0.5\columnwidth]{images/sample_snapshot_of_simulation.eps} }
 \caption{ Snapshots of two MS systems: a collagen fiber structure with 890,000 atoms (top) and a dipalmitoylphosphatidylcholine (DPPC) bi-layer lipid system with 402,400 atoms (bottom) \cite{mainArticle}}
 \label{fg:collagen_dipal}
\end{figure}

\section{Problem Statement}
Simulation software systems, in general, follow the same methodology of running and storing simulation data. The simulation software system examples are: Gromacs \cite{GROMACS4}, VMD\cite{VMD}, MDAnalysis~\cite{MDAnalysis}, Wordom~\cite{wordom}, MD-TRACKS~\cite{MDtracks}, SimulaidOne~\cite{Simulaid}, Charmm~\cite{CHARMM}. In the type of simulations brought up as examples above, the flow of the data is the following. Once the simulation is run, the output files are contained as trajectory files with descriptors (they contain information about space dimensions, number of atoms and frames, etc.) that can be easily transposed into simple flat files containing the physical properties atom by atom, frame by frame, which are consequently read and processed by the proposed push-based system. Since we have certain amount of queries needed to be run on given simulation data, generating high I/O traffic followed by design of pull-based system is not considerable. The approach proposed in the previous paper is very good in terms of performance in comparison with the original pull-based system\cite{mainPaper}. As you can see on Figure \ref{fg:sample_estimation_old_paper}, it gets up to over 100 times of speedup which is quite crucial, if we think about days of computations. The problem is still that some of the queries processed by those means are still improvable, especially taking into consideration the fact that in the used previous works for calculation of 2-body functions, which take the biggest time for processing~\cite{ytu:icde09, EDBT12}, we might have some error bounds, which might be unacceptable for certain simulation analysis where sacrifice on loss of data is unbearable. Besides it, although we might have huge performance boost on specific data structure as density map, we still have certain expectation from data nature (such as its uniformity), thus, it makes sense to have desire to simplify the processing and still having pure computation based on any values, considering that the queries should be available to be pre-programmed by user in a separate module of the code. All the benchmarks will be done in comparison with already push-based system versus push-based system with GPU improvements. We think that it is  not fair to compare pull-based against GPU push-based approach, since the goal of the paper is to demonstrate capabilities of GPU approach that can beat simple sequential push-based system.

\begin{figure}
 \centerline{ \includegraphics[width=0.5\columnwidth]{images/speedup838K-pomalo-za-4.eps} }
 \caption{ Speed up over different levels of atom selection for 838,400 atoms \cite{mainPaper}}
 \label{fg:sample_estimation_old_paper}
\end{figure}

\section{Our approach with improvement}

Reviewing the approach we are coming up with and what it was before helps us analyzing the weak parts and things that we can focus on to eliminate old bottlenecks.

Original non-improvement standard approach with pull-based querying is probably one of the most flexible, because nobody has to think about improvements and complexities, it does it for every chunk of data and pulls it whenever it is needed. As it has already been mentioned, at this point, the biggest challenge is that supposing we have $M$ queries, for $N$ atoms we would have to load every chunk $M$ times for $N$ chunks of data (frames), and apparently it is very time consuming of I/O and non-linear computations, and that is only for linear functions. Improvements could be done on caching level, but again the worst case would be the same complexity, because most of the time there is nothing really to cache, and if there is, it is too big to store it in memory, so, it just has to be re-read from file. Using accumulative data structures and trying to optimize by means of them doesn't help either, because, first of all, it is too big data to store in memory, and, secondly, most of the time it is based on data distribution expectation of some kind, which is not the best solution, because the experiments can be different, thus data distribution is sometimes not predictable (boiling down to a specific physics problem will destroy the abstractness and applicability of the system to variety of other problems and will narrow it down making it less useful). In summary, the bottlenecks in this approach are I/O time consumption and slow data processing (especially for non-linear computation).

After all, push-based system for solution of the given problem is tremendously fast in comparison with pull-based system. General idea is to push a chunk of data (frame) through the pre-defined network of queries that will be needed to run on the data sets, which basically means chunk of data will be read from file and loaded into memory only once in order to go through all the queries. As you can observe \cite{mainPaper}, it can speed up for more than 100 times (depending on workload, number of atoms, and functions being used). It is clear that the biggest challenges in this approach were related to implementation of non-linear functions like SDH, since once the data is pushed through the queries, it is almost not coming back (if it is coming back, the point of optimization is lost), and the idea is to come up with approach to eliminate this overhead and weakness. In this case, it is also important what memory exactly is being used, because trying not to use used parts of the same chunk also matters, because L-level caches of CPU might play a huge role in big data sets. One of the solutions is building quad-tree based data structure that will help optimize the calculation, and it has some expectation from data distribution which can speed up depending on dimensions (2D and 3D). This is a mutual problem of both data loading (memory usage) and its processing, and push-based system is a great approach, because it still remains as an abstract concept applicable to different queries and probably wide variety of other problems of similar kind in different fields.

Finally, we've analyzed some main ideas, advantages, and disadvantages of our approach in this paper which is basically push-based system with optimization on data processing part by means of GPU computation (in other words, parallelism). Reading the data from files is a huge overhead, thus, speedup of push-based system in general is significant over pull-based system in our given case. 
It's been demonstrated in the previous paper with different amount of atoms, frames, and workloads. Since we have to load the data every time there is a different query versus push-based system loads a chunk of data once, it is quite noticeable how the framework contributes to efficiency of analysis of simulation data. You can observe some estimation samples in Figure \ref{fg:sample_estimation_old_paper}. Even though the memory loading and pushing steps are the key features of the proposed design, nevertheless, having general knowledge about the network of queries, in this paper, we will try to focus on optimizing the actual execution of them.

Some of the queries may be very slow due to their nature on a sequential type of computation,  especially, 2-body functions. Since the nature of the data that comes with simulation is basically physical properties of atoms, there is a lot of computation that involves independent primitive mathematical operations, which makes it a perfect problem for parallelism. For example, image processing computations and graphics processing are done on GPUs, because they have a lot of minor primitive independent pixel-based computations.

As it has already been mentioned, although the original idea of push-based system for molecular simulation data analysis was focused on optimizing throughput and usage of loaded data, there were some extra approaches specifically for 2-body functions. For example, Density Map for Spacial Distance Histogram was used in order to avoid additional memory allocation and latency reduction with proven error bounds. SDH is a quite intensive and computational problem, especially with increasing amount of atoms.

@@

\section{Contribution and roadmap of the paper}

In collaboration with Physics Department, the original motivation came from the need of improving analytic query executions performance over big data being generated by simulation software. The idea was to take input stream coming from simulation software (ideally live) and run network of queries very efficiently, because sometimes they might take days and more to compute. The proposed improved version of push-based system for molecular simulation data analysis, we believe, gives an opportunity for scientists to run their analysis fast having an ability to expand network of queries and customize computation functions to get the desired results. 
Our system primarily represents a system that works in several modules that were developed in C, CUDA, Python programming languages that, in combination, give us the desired result to perform efficient computations for the network of queries developed in the previous paper \cite{mainPaper} as well as to prove the concept of optimization of push-based system by means of parallelism with no data loss. We've developed the following modules:

\begin{itemize}
\item[1.] Mock data streamer and generator - tool developed in Python programming language that helps to run performance tests (could also be of other simplest tests like regression) on any kind of data and can be used to simulate random values for physical properties of atoms. The simplest usage of the tool is running a python script with arguments (number of atoms and number of frames).
\item[2.] Main execution script - the main module developed in C and CUDA for the actual computation of the network of queries. Network of queries comes predefined and has to be compiled in the script (though later can be abstracted as a configuration file or argument for interpreting general functions), along with computation implementation function that interact with a given GPU.
\item[3.] Data conversion - this module comes as a part of the code in Main execution script for the users to be able to manipulate data conversion before loading it to GPU. This is for advanced users having capability of optimizing computations and queries on any level of memory hierarchy (CPU and GPU). It is an abstract part which has to be used upon specific cases to optimize specific functions, since there is no way to predict what functions need to be run and how they can be optimized.
\item[4.] Sequential CPU based Push-based system - we re-created original push-based system with the same network of queries as a benchmark for estimation of speedup results of the new approach

\end{itemize}

Taking an advantage of GPU devices nature and incorporating parallelism and streaming for different types of queries, we have come up with a good speed up over sequential processing which gets up to 60 times faster, which is just for sampling with no deep improvements on individual queries (since we believe that it should represent average user usage). Original push-based system already had a good speed up in terms of performance in comparison with original pull-based approach and floats around more than 100 times faster depending on number of atoms, frames, and workload. The module for generation of mock data in Python  was used for benchmarks, and it represents realistic and pessimistic approach of estimation, since there is no data distribution for physical properties values. We believe that it might be quite important, because experiments may vary widely. This mock data has exactly the same information that would come with Gromacs simulation files, and it has exactly the same format as in the previous paper \cite{mainPaper} right before loading it to memory. This improved version has been developed on Amazon Elastic GPU \cite{amazonGPU} nodes that consequently can be replicated and scaled up becoming more of a streaming distributed processing engine (which can be very effective on costs, since it is always easy to scale them up, shut them down and spin them back up), as well as on a simple desktop computer with a GPU device.
In general, major contribution of this paper is the framework of push-based system with an incorporated parallelism based on CUDA programming language. This tool can be integrated into open source project of Gromacs as a custom plugin that can be used for optimized analytic queries to be run over data sets live streaming as well as from files. This will be discussed in more details in further chapters.
The structure of the framework which will be described later is developed in such a way that one can easily add new queries based on their complexity or modify existing ones adding appropriate selections and filters.

@@

\chapter{Related Work}\label{sc:relatedwork}

Push-based System for molecular simulation data analysis with its improved GPU features has couple of main focuses in terms of technical implementation and challenges related to those that have already been tried to be solved by other different approaches and software. Here is the review of other works.

\section{Scientific software}

First of all, this system is a tool for scientific computation and in this case particularly for molecular simulation data analysis. PeriScope is a project and initiative by University of Michigan which is a system that will optimize data management for scientific explorations. Relational Database Management Systems were created to optimize business applications and it was revolutionary. Those couldn't be built without these systems with abstract languages like SQL, since it manipulates and automatically takes use of indexes, etc. PeriScope is a project that tries to come up with same approach against scientific data and computation with respectively larger scale.\cite{PeriScope} It is kind of difficult to compare our proposed system with PeriScope, but the original motivation is very similar: optimize scientific computation. PeriScope is a larger project directed towards more abstract solution like RDBMs that would work for any type of data. Another existing tool for scientific data and optimized processing is SciDB. It was created to make scientific data storage and querying better and faster. It is very similar to traditional RDBMS systems in terms of querying language and represent data storage in tables, but states to be faster than Postgres in 2 orders of magnitude. Some of the main features is distributed processing with atomicity and durability (the latter is quite difficult for large scale data), while the disadvantages are no-overwriting (only append mode), single value columns only (no arrays, etc), where records are like simple tuples with join function on querying multiple tables \cite{SciDB}. One of the major differences between Push-based Molecular simulation data analysis tool and the projects above is that our proposed design doesn't do anything with storing the data, because we are mainly concentrating it as a side plugin for processing and querying data in an optimized way, while the actual data generation and storage will be done with the actual simulation software.

\section{Stream processing}
Another major modern issue is live data streaming and live processing, which is also possible with our system. Basically, streaming function is a big challenge and need nowadays, because such networks as Twitter and Facebook are able to keep track of live trends and events, including people's reaction and feedback that have to be reflected real time for tremendous amount of users across the globe. Borealis Stream Processing Software is a tool being developed by MIT, Brown, and Brandeis Universities. It's primary goal is a distributed stream processing and based on Medusa and Aurora. Some of the major functions of the software are highly available distributed large streams processing live, dynamic and highly-scalable with possibility of changing queries live and dynamic vision of query results. \cite{borealis} There is also a similar need in scientific world, since being able to keep track of results in real time might be crucial to be able to make certain explorations.

\section{Big Data Frameworks}

We are potentially dealing with data that could easily get up to terabytes and petabytes in sizes, and it will be fair to apply term of big data. Nowadays there are bunch of frameworks that try to solve this problem. Some of these frameworks are Hadoop \cite{hadoop}, Hadoop with GPU extension \cite{hadoopgpu}, Spark \cite{spark}, Flink \cite{flink} (this is more of a big data stream processing framework) and other ones that are possibly just custom modifications and distributions of these. The mentioned tools help dealing with big data giving an ability to process them in batches on clusters using underlying HDFS (Highly Distributed File System), and normally go along with other utilities that help to apply MapReduce concept. In fact, the frameworks listed above could even be use for our problem on top of our system, because, as we discuss in section \ref{sc:conclusion}, it could be extended to a distributed system for analyzing independent frames in parallel.

@@

\vspace{3mm}
\chapter{The system design}\label{sc:system}
\vspace{3mm}

@@

\section{Data flow architecture}
